---
title: "Lab 18"
author: "Lisa Chen A17082974"
format: html
editor: visual
---

## 

First we will examine and explore Pertussis case numbers in the US as tracked by the CDC

https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html>

We can use the datapasta package to scrape this data from the website into R


>Q1 Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         No.Reported.Pertussis.Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )

View(cdc)
```


```{r}
head(cdc)
```


I want a plot of cases per year with ggplot


```{r}
library(ggplot2)
```
 >Q2  Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
aes(year, No.Reported.Pertussis.Cases) +
  geom_line() +
  geom_vline(xintercept = 1947, col="blue") +
  geom_vline(xintercept = 1992, col="red") +
  geom_vline(xintercept = 2020, col="orange")
  
```


>Q3  Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There is an increase most likely due to COVID and comorbidities that could contribute to decreased immunity. THere could also be 1) more sensitive PCR-based testing, 2) vaccination hesitancy 3) bacterial evolution (escape from vaccine immunity), 4) waning of immunity in adolescents originally primed as infants with the newer aP vaccine as compared to the older wP vaccine.



Access data from the CMI-PB project

This database(like many modern project) uses an APi to return JSOn format data. 

We will use the R package jsonlite
```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/subject", 
                     simplifyVector = TRUE)
head(subject)
```
> q4How many wP(the older whole-cell vaccine) individuals and aP (newer acellular vaccine) individuals are in this dataset?



```{r}
table(subject$infancy_vac)

```

> Q5. What is the number of individuals by biological sex and race

```{r}
table(subject$biological_sex)
```
> Q6 What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)

```{r}
table(subject$race, subject$biological_sex )
```
```{r}
subject$year_of_birth
```
#Side-Note: Working with dates

We can use the libridate package to ease the pain of doing math with dates. 


>Q7 Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}

library(lubridate)
```

```{r}
today()
```
```{r}
today() - ymd("2002-01-01")
```

```{r}
today() - mdy("5-15-2002")
```





>Q8 Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```


So what is the age of everyone on our dataset. 

```{r}
time_length(today() - mdy("5-15-2002"), "years")
```

```{r}
subject$age <- time_length(today()-ymd(subject$year_of_birth), "years")
```



>Q9  With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?



```{r}
library(ggplot2)
ggplot(subject) +
aes(age) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  geom_histogram()

```

##Get more data from CMi-PB



```{r}

specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector=T)
head(specimen)
```

Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:


we need to **join** these two tables(subject and specimen) to make a single new "meta" table with all of our metadata
```{r}
library(dplyr)

meta <- inner_join(subject, specimen)

head(meta)


```
Now we can read some of the other data from CMI-PB

```{r}
ab_titer <- read_json("http://cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector=TRUE)

head(ab_titer)
```

One more 'inner_join()' to add all our metadata in 'meta' on to our 'ab_data' table: 


>Q.10 

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```
> Q11

```{r}

table(abdata$isotype)
```



> Q12

```{r}
table(abdata$antigen)
```
```{r}
ggplot(abdata) + 
  aes(MFI, antigen) +
geom_boxplot()
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

why are certain antigens and not others very variable in their detection levels here?

can you facet or even just color by infancy_vac? Is there some difference?
```{r}
ggplot(abdata) +
aes(MFI, antigen, col=infancy_vac) +
geom_boxplot()
```
>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

You can use the CMI-PB website search functionality and Terminology Browser (under development) to find out about each antigen. Note that this is still work in progress.



There are potentially some differences here but in general it is hard to tell with this whole dataset overview...

```{r}
table(abdata$dataset)
```


Lets focus in on just the 2021 dataset
```{r}
abdata.21 <- filter(abdata, dataset == "2021_dataset")
table(abdata$dataset)
```


Focus on PT antigen IgG levels

```{r}
pt.21 <- filter(abdata.21, isotype == "IgG",  antigen == "PT")

```

plot of days(time) relative to boost vs MFI leves

>Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
ggplot(pt.21)+ 
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() 
```
> Q16. What do you notice about these two antigens time courses and the PT data in particular?

PT levels clearly rise over time and far exceed those of OVA. They also appear to peak at visit 5 and then decline. This trend appears similar for wP and aP subjects.

> Q17. Do you see any clear difference in aP vs. wP responses?

Yes I do see differences in an increased MFI values in the beginning planned day to relative boost.
```

